---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.


### 1. Loading and preprocessing the data
```{r, loaddata, warning = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)

## Load data
activitydata <- read.csv("activity.csv", stringsAsFactors = FALSE)
activitydata$date <- as.Date(activitydata$date, "%Y-%m-%d")

## Show the first six lines of the data
head(activitydata)
```

### 2. What is mean total number of steps taken per day?
```{r, dailysteps}
## Calculate total number of steps taken per day
dailysteps <- tapply(activitydata$steps, activitydata$date, sum, na.rm = TRUE)
dailysteps

## Histogram of total steps per day
qplot(dailysteps, geom = "histogram", binwidth = 5000, 
      main = "Histogram of total steps per day", 
      xlab = "Steps per day", ylab = "Count") +
        theme_bw(base_size = 12)

## Calculate mean and median of the total number of steps take per day
dailysteps_mean <- round(mean(dailysteps), digits =0)
dailysteps_median <- median(dailysteps)
```

**The mean of the total numbers of steps taken per day is `r dailysteps_mean`. The median of the total numbers of steps taken per day is `r dailysteps_median`.**


### 3. What is the average daily activity pattern?
```{r, intervalaverage}
## Calculate the average number of steps per interval
intervalaverage <- activitydata %>%
        group_by(interval) %>% summarise(steps = mean(steps, na.rm = TRUE))

## Time series plot
qplot(interval, steps, data = intervalaverage, geom = "line", 
      xlab= "Interval", ylab = "Average numbers of steps",
      main = "Time series plot of average steps over 5-minutes intervals") +
        theme_bw(base_size = 12)

## Which interval has the highest avearge numbers of steps
interval_highest <- intervalaverage[intervalaverage$steps == max(intervalaverage$step), 1]
interval_higheststeps <- round(intervalaverage[intervalaverage$steps == max(intervalaverage$step), 2], digits = 0)
```
**The interval with the highest average number of steps is the interval `r interval_highest`, which has average `r interval_higheststeps` steps.**


### 4. Imputing missing values
```{r, imputemissing}
options(scipen=8)
## Calculate total number of missing data
missing <- sum(is.na(activitydata$steps))
print(paste("There are total", missing, "missing data"))

## Split the original dataset into the good data and the missing data 
index <- is.na(activitydata$steps)
gooddata <- activitydata[!index,]
baddata <- activitydata[index,]

## Replace the missing steps with the average steps of that interval
for(i in 1:nrow(baddata)){
        whichinterval <- which(intervalaverage$interval == baddata[i, 3])
        ## the missing value is replaced by the rounded number
        baddata[i, 1] = round(intervalaverage[whichinterval,2],digits = 0)
}
## Create a new dataset to fill in the missing value with average steps of that interval
newdata <- rbind(gooddata, baddata)

## Calculate total number of steps taken per day for newdata
newdailysteps <- tapply(newdata$steps, newdata$date, sum)
newdailysteps

## Histogram of total steps per day
qplot(newdailysteps, geom = "histogram", binwidth = 5000, 
      main = "Histogram of total steps per day", 
      xlab = "Steps per day", ylab = "Count") +
        theme_bw(base_size = 12)

## Calculate mean and median of the total number of steps take per day
newdailysteps_mean <- round(mean(newdailysteps), digits =0)
newdailysteps_median <- median(newdailysteps)
```
**The missing data of the steps in the original dataset is filled by the average steps of the interval where the missing occurs. Imputing missing data changed the distribution of total steps per day. Now the mean of the total numbers of steps taken per day is `r newdailysteps_mean`. The median of the total numbers of steps taken per day is `r newdailysteps_median`.**


### 5. Are there differences in activity patterns between weekdays and weekends?
```{r, wkday}
## Create a new variable with 2 levels "weekday" & "weekend"
weekday <- c("Monday", "Tuesday","Wednesday","Thursday", "Friday")
newdata2 <- mutate(newdata, 
                   day = ifelse(weekdays(date) %in% weekday, 
                                "weekday", "weekend"))

## Calculate the average number of steps per interval
wkdayaverage <- newdata2 %>%
        group_by(day, interval) %>% summarise(steps = mean(steps))


## Time series plot
qplot(interval, steps, data = wkdayaverage, geom = "line", 
      facets = day~.,
      xlab= "Interval", ylab = "Average numbers of steps",
      main = "Time series plot of average steps over 5-minutes intervals") +
        theme_bw(base_size = 12)

```

**As shown in the plot, the time series pattern of numbers of steps are different in weekdays and weekends.**
